### 2020-02-23 rebuttal/rebuttal.tex

\documentclass[11pt,letterpaper]{article}
\usepackage[top=0.85in,left=0.85in,footskip=0.75in]{geometry}
\usepackage{mathpazo}
\usepackage{mdframed}
\usepackage{graphicx}
\usepackage{sectsty}
\allsectionsfont{\normalfont\sffamily\bfseries}
\subsectionfont{\normalfont\sffamily\bfseries}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{amsmath}

% To place figures exactly on a page
\usepackage{float}
\usepackage{graphicx}

% Bold the 'Figure #' in the caption and separate it from the title caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,font=small,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

% Strike out
\usepackage[normalem]{ulem}

% Custom BiBTeX style
\bibliographystyle{bibStyle}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{helvet}

% To define more useful LaTeX commands
\usepackage{xparse}
\usepackage{graphicx}

% Equation referencing
\DeclareDocumentCommand \eref{oooo} {\IfNoValueTF{#2}{Eq. R~(\ref{#1})}{\IfNoValueTF{#3}{Eqs.~(\ref{#1}) and (\ref{#2})}{\IfNoValueTF{#4}{Eqs.~(\ref{#1})-(\ref{#3})}{Eqs.~(\ref{#1})-(\ref{#4})}}}}

% Figure referencing
\DeclareDocumentCommand \fref{ooo} {\IfNoValueTF{#2}{Fig.R~\ref{#1}}{\IfNoValueTF{#3}{Figs.~\ref{#1} and \ref{#2}}{Figs.~\ref{#1}-\ref{#3}}}}
\renewcommand{\thefigure}{R\arabic{figure}} 
\renewcommand{\theequation}{\sffamily R\arabic{equation}} 
\renewcommand{\thetable}{\sffamily R\arabic{table}}

%% setup the response environment
\mdfdefinestyle{revstyle}{
outerlinewidth=2pt,
bottomline=true,
leftline=true,rightline=true,
topline=true,
font=\small,
skipabove=\baselineskip,
skipbelow=\baselineskip,
roundcorner=10pt,
innertopmargin=0.5em,
innerbottommargin=1em,
splittopskip=1em,
splitbottomskip=1em
}

\mdfdefinestyle{respstyle}{
outerlinewidth=2pt,
bottomline=false,
leftline=false,rightline=false,
topline=false,
font=\small,
backgroundcolor=gray!20,
skipabove=\baselineskip,
skipbelow=\baselineskip,
roundcorner=10pt,
innertopmargin=0.5em,
innerbottommargin=1em,
splittopskip=1em,
splitbottomskip=1em
}

\newmdenv[style=revstyle,frametitle=\sffamily \itshape Reviewer Comment]{review}
\newmdenv[style=respstyle,frametitle=\sffamily \itshape Author Response]{response}
\newmdenv[style=respstyle,frametitle=\sffamily \itshape Author Response: Executive Summary]{responseSummary}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% END MACROS SECTION%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\section*{Response to Reviewer for PONE-D-19-32846 ``Theoretical Investigation of a Genetic Switch for Metabolic Adaptation''}
Kathrin S. Laxhuber,$^{1}$ Muir J. Morrison,$^{2}$ Griffin Chure,$^{3}$ Rob Phillips$^{2, 3, \dagger}$
\vspace{10pt}

\noindent\tiny{$^1$ Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland,
$^2$ Department of Physics, California Institute of Technology, Pasadena, CA, USA,
$^3$ Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA, $^\dagger$ E-mail: \texttt{phillips@pboc.caltech.edu}}

\begin{responseSummary}
We thank the editor and reviewer for considering our manuscript and
bringing some important points to our attention. The reviewer's comments
significantly improved our work and inspired us to make very interesting
comparisons between the deterministic and stochastic picture of our
system. Furthermore, as rightly suggested by the reviewer, we have now
added experimental data to the manuscript and expanded the author list
to reflect this. Below, we address the reviewer's concerns one-by-one
and indicate where changes were made either to the main text or the
supporting information. Besides these changes, we made a few small
corrections to the text that had slipped our attention before. We
believe we have resolved the three issues brought up by the reviewer and
hope our manuscript is now ready for publication in PLOS ONE.
\end{responseSummary}

\subsection*{Reviewer \#1}
\begin{review}
In their manuscript "Theoretical investigation of a genetic switch for
metabolic adaptation", Laxhuber and colleagues present a very nice
theoretical analysis of the behavior of the E. coli xapABR system, using
both deterministic and stochastic simulations to demonstrate that the
system can show meaningful and important bistability driven by the
activity of the membrane transporter. In general the work is well done,
and it is technically of high quality, but at present I have two
substantial concerns that I think must be addressed in order for the
paper to live up to its potential.
\end{review}

\begin{response}
We thank the reviewer for their careful reading of our work and are
pleased that their impression of the manuscript is favorable overall.
\end{response}

\begin{review}
My first concern is that while the authors present their deterministic
simulations first, and draw many of their conclusions primarily from
those simulations, it seems to me that the use of such simulations is
contra-indicated by the authors' own findings. As both the number of
XapR molecules and mRNA molecules are small integers (likely 10 or
less), treatment of these species as continuous, deterministic
quantities can never be more than a poor approximation. I find this
troublesome for a couple of reasons. First, the authors still treat
their continuous treatment as an apparently 'good' way to model a
genetic switch like that at work at the xapAB promoter; while I find
their analysis insightful, in the end I cannot support such a
recommendation given the stochasticity intrinsic in the low protein and
mRNA copy numbers present here.
\end{review}

\begin{response}
We agree with the concern of the reviewer that in general, deterministic
approximations of systems with such low copy numbers should not be
trusted. We do not intend to give the general recommendation that a
continuous treatment alone is usually sufficient, and in order to avoid
this, we added some words of warning in the beginning of our
deterministic as well as our stochastic analysis.

Nevertheless, despite the low copy numbers,
for the system we investigate the deterministic model is a
surprisingly accurate description of the means of copy number and
waiting time distributions produced by the stochastic simulations.
To make this more apparent, we added some additional comparisons of the two
models. We address these in more detail in our responses below,
but to summarize: we added a bifurcation diagram
(Fig~\ref{bifucation}) which illustrates that the stochastic
distributions track the deterministic fixed points surprisingly well,
and we compared the stochastic timescales with the solution of the
deterministic model (Fig~\ref{adaptationtime}) and found excellent
agreement there too.
Even though stochastic fluctuations around the deterministic
trajectories may be large, the stochastic trajectories do still flow to
the fixed points as predicted by the deterministic model, except when
the system is close to either of the bifurcation points in
Fig~\ref{bifucation}, where it is unsurprising that fluctuations are
magnified in importance.
Hence, while we find the close agreement surprising, we believe it to be
appropriate to show and draw conclusions from deterministic phase
diagrams. The latter have the advantage of providing the reader with a
more intuitive understanding of the system than the plots of the results
from the stochastic simulation.
\end{response}

\begin{review}
Second, while the authors do technically recapitulate the bistable
behavior of the system in the stochastic version, the bistability seems
based on their description to be quite fragile and not very meaningful,
since it only occurs here for small distances between the lower and
upper fixed points
\end{review}

\begin{response}
We thank the reviewer for bringing this point to our attention, which we
believe to be a misunderstanding that resulted from a poor explanation
in our original text. We believe we have clarified this in the
manuscript, and allow us elaborate here too: Stochastic bimodality occurs
only when two of the three fixed points are close together, or in other
words, when the system is poised ``sufficiently near'' one of the
bifurcation points. This may become clearer when looking at the
bifurcation diagram that we added to our section about hysteresis at the
end of the main text. For convenience, we insert this plot below in
Fig~\ref{bifucation} which addresses what we believe the reviewer is
referring to: only when two of the three fixed points are close do the
stochastic trajectories occupy both fixed points, resulting in
bimodality.

Since this bimodality only occurs for a small region in parameter space,
this actually indicates strong and not fragile bistability. In other
words, the fact that the system does not stochastically ``jump'' to the
other fixed point means that the fixed points are very stable to
stochastic fluctuations. Thus, the switch-like feature is robust:
there are two distinct states that the system can occupy, and only near
the bifurcation points is it likely for stochastic trajectories to
transition between them. Put differently, deterministic bistability does
not need to result in stochastic bimodality, and in fact stochastic
bimodality is contrary to robust bistability.
\end{response}

\begin{figure}[h!]
	\centering
	\includegraphics{Fig10_bifurcation.eps}
	\caption{{\bf Bifurcation diagram showing the hysteresis loop in the stochastic system.}
		The blue line shows the positions of the deterministic fixed points, while the dashed line indicates the instability of the middle one. The orange and yellow points show the mean of the stochastic simulations when started at zero and at the high fixed point, respectively. The positions of the two peaks in the bimodal distributions are indicated by smaller points, connected by dotted lines. The arrows illustrate the hysteresis loop in the stochastic simulations.}
	\label{bifucation}
\end{figure}

\begin{review}
	(and seems to take an unreasonably long time to switch between states).
\end{review}

\begin{response}
To address the indeed valid point of unreasonable switching times, we
have compared the timescales in our stochastic simulations with those from the
numerical solution of the deterministic model. We find excellent
agreement, as shown in an updated figure in the main text and also below
in Fig~\ref{adaptationtime}. So, again perhaps surprisingly,
stochasticity does not materially alter the mean switching time as
computed from the deterministic picture. Even adding additional
stochasticity through transcriptional bursting (discussed more below)
does not alter the mean switching time, though it does broaden the
distribution of times.
It is possible that this broadening resolves the apparent paradox. Even
if the mean switching time is quite long, a non-negligible fraction of a
population of cells could nevertheless switch on a shorter time-scale,
especially with inclusion of extra stochasticity beyond what we have
explicitly considered.
\end{response}

	\begin{figure}[h!]
		\centering
		\includegraphics{Fig9_evolution.eps}
		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
			(A) Blue shows the results from one run of the stochastic simulation, orange shows the trajectory obtained from solving the deterministic ODE's. The simulation was started
			at an mRNA, protein and intracellular xanthosine count of 0. The
			extracellular xanthosine concentration was chosen to be
			$\mathrm{[c]_a} = 25$. (B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 1000 runs of the simulation (same conditions as (A)). The orange line shows the corresponding deterministic time. The green dashed line right next to it indicated the mean of the blue distribution.}
		\label{adaptationtime}
	\end{figure}

\begin{review}
I think the authors are likely correct in their assertion that they
underestimate the amount of noise in the system due to not including
transcriptional and translational bursting, and that the transition
rates would likely be faster, but this gives no idea of how *much*
faster, and thus it seems premature to simply declare victory and move
on.
\end{review}

\begin{response}
This is a very reasonable concern which we have investigated more
thoroughly now. We added rather strong transcriptional bursting to our
simulation and do not find qualitative changes in the results. In
particular, fluctuations around the mean become larger and, accordingly,
distributions broader. Furthermore, bimodality begins to appear for
lower values of the extracellular xanthosine concentration. Very
interestingly, the change in the \emph{mean} transition time between the
stable fixed points is negligible. However, the \emph{variance} in this
time across many runs of the simulation becomes larger and the
\emph{peak} of the distribution shifts to shorter times. We have
included two new figures in the SI discussing these results (Figs. 6 and
7), along with discussion there and in the main text.
%	Below, we also show the two new figures from S1 Text that illustrate the similarities and differences between the simulation with and without transcriptional bursting.
\end{response}

%\begin{figure}
%	\centering
%	\includegraphics[width=1\textwidth]{FigSI6_7.pdf}
%	\caption{{\bf Time evolution and adaptation times from the stochastic simulation when transcriptional bursting is included.}
%		(A) shows the time evolution from one run of the simulation. The simulation was started
%		at an mRNA, protein and intracellular xanthosine count of 0. The
%		extracellular xanthosine concentration was chosen to be
%		$\n{[c]_a} = 25$. (B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 1000 runs of the simulation (same conditions as (A)). The distribution from the main text (without bursting) is shown in orange for comparison.}
%	\label{evolbursts}
%\end{figure}
%
%\begin{figure}
%	\centering
%	\includegraphics[width=1\textwidth]{FigSI6_7_2.pdf}
%	\caption{{\bf Distributions from the stochastic simulation when transcriptional bursting is included.}
%		Blue shows the distributions from the simulation with bursts, orange shows those without bursts from the main text for comparison. The following values of $\mathrm{[c]_a}$ were used: (A) $\n{[c]_a}=12$, (B) $\n{[c]_a}=17.5$ (not shown in main text), (C) $\mathrm{[c]_a}=18.5$, and (D) $\mathrm{[c]_a}=25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
%		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$). All other parameters are as in the table in the main text, and the simulation
%		was run 1000 times for $10^6 \unit{s}$ each (simulated time) and
%		started at a mRNA, protein and intracellular xanthosine count of 0.}
%	\label{distbursts}
%\end{figure}

\begin{review}
My recommendation would be that the authors put in some more cautionary words about the applicability of the continuous model, and ideally spend a little more time characterizing the discrete model. in terms of switching times, parameters that give rise to bimodal behavior, effects of incorporating reasonable amounts of transcriptional bursting, etc. Some more direct comparisons of the behavior of the two models in similar ranges of parameters space would also likely be enlightening.
\end{review}

\begin{response}
We hope that we have addressed all these points in our responses above
and shortly summarize here: We put in some cautionary words about the
general usage of continuous models and demonstrated that for our system
and the conclusions we draw, the continuous model is surprisingly appropriate.
Furthermore, we extended our characterization of the stochastic model as
suggested by investigating the distribution of adaptation/switching
times as well as by incorporating transcriptional bursting in our
simulation. The deterministic and stochastic model are now directly
compared (for the same set of parameters) in the bifurcation diagram as
well as in terms of their adaptation/switching times.
\end{response}

\begin{review}
My other major difficulty here is that essentially all of the work is
premised on reference 22, which is 'unpublished data.' This is the only
experimental evidence given for the bistability of xapAB expression,
justification of many of the model parameters, experimentally expected
switching timescales, etc. It is impossible to properly interpret the
present work, either for the purposes of review or from the viewpoint of
an interested reader, without availability of those data, which either
ought to be their own independent publication or ought to be included
with this one.
\end{review}

\begin{response}
With further consideration, we agree with the reviewer that we relied
rather heavily on this unpublished data and that it is therefore
essential to make the experimental data available.
\textbf{Also address changes to author list once finalized.}
We have opted to include the data with this publication rather than
separately. We have added a new section, ``experimental motivation,'' in
the beginning of the main text which discusses the main features of the
data that are relevant for our model. Furthermore, we added a section on
the materials and experimental methods at the end of S1 Text (section
E).
The raw data is
publicly available at \textbf{!!still need to sort out
DOI/caltechData/ETHZ hosting/...??}.
\end{response}

%\begin{figure}[h]
%    \centering{
    %\includegraphics{figure_path}
%     \caption{\textbf{Figure caption.}}
%    \label{fig:figlabel}}
% \end{figure}

\bibliography{library}

\end{document}

### 2020-02-23 main_text/main_text.tex
Add rebuttal letter + minor associated edits to main text

	The results agree very well with the deterministic fixed points and
	experiments: the mean numbers of mRNA, protein, and xanthosine in the
	stochastic results are as predicted from the phase portraits. It does,
	however, become clear that the phase portraits do not tell whether the cells
	will actually populate both the high and the low expression state,
	because they do not show the effective barrier height between the two states.
	In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
	expression state during the run time of our simulations. This actually is a good finding: it implies that bistability, characterized by two separate \emph{stable} steady-states, largely persists even in the presence of stochastic fluctuations. Thus, the fact that bimodality only occurs in a small regime means that the circuit is a strong switch and the deterministic picture is sufficient except in a small range of parameter space. Whether or not that is where the system actually is remains a question for future experiments.

	We found that the two lower fixed points (marked as 1 and 2 in
	Fig~\ref{fig4:bistability}) need to be very close like in
	Fig~\subref{fig8:stochC}{(B)} to give bimodality. For lower $\n{[c]_a}$,
	meaning larger distance between the first and second fixed point,
	almost no switching was observed. Of course, switching is also a matter of the
	waiting time and stochastic effects: if one waits for long enough, it
	should eventually occur. However, switching times of more than several hours
	are not at the center of this investigation and would mean that switching is
	extremely unlikely. There are two aspects that become relevant in this
	context that we neglect in our analysis but briefly mention here:
	transcription and translation bursts lead to higher stochasticity and cell
	division leads to some discontinuity in the process. The effect of bursts is addressed in~\nameref{S1_Text}.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig9_evolution.eps}
		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
			(A) Blue shows the results from one run of the stochastic simulation, orange shows the trajectory obtained from solving the deterministic ODE's. The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
			at an mRNA, protein and intracellular xanthosine count of 0. The parameters that were used are the same as in
			Table~\ref{table1:nondim}, the only exception being the
			extracellular xanthosine concentration, which was chosen to be
			$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
			$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$) just as in
			Fig~\subref{fig8:stochC}{(C)}.
			(B) In blue we plot the waiting time distribution for 1000
			runs of the simulation (same conditions as (A)) to reach
			the steady-state mean.
			We define this time as the elapsed time when the protein
			copy number first reaches 90\% of its value at the upper
			fixed point.
			To better visualize the bulk of the distribution, we
			excluded from the plot $\sim10$ runs with adaptation times
			larger than $6 \cdot 10^5 \unit{s}$.
			The green dashed line indicates the mean of the blue distribution.
			The orange line shows the corresponding deterministic time.
			}
		\label{fig9:StochT}
	\end{figure}

>>>>

	The results agree very well with the deterministic fixed points and
	experiments: the mean numbers of mRNA, protein, and xanthosine in the
	stochastic results are as predicted from the phase portraits. It does,
	however, become clear that the phase portraits do not tell whether the cells
	will actually populate both the high and the low expression state,
	because they do not show the effective barrier height between the two states.
	In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
	expression state during the run time of our simulations. This result
	implies that bistability, characterized by two separate
	\emph{stable} steady-states, largely persists even in the presence
	of stochastic fluctuations, suggesting that the deterministic
	picture is surprisingly and unreasonably effective. In other words,
	the fact that bimodality only occurs with some fine tuning of
	parameters means that the circuit is a strong switch and the
	deterministic picture is sufficient except in a small region of
	parameter space. Whether or not the system is actually in that
	region of parameter space remains a question for future experiments.

	To elaborate on the preceeding statements,
	we found that the two lower fixed points (marked as 1 and 2 in
	Fig~\ref{fig4:bistability}) need to be very close like in
	Fig~\subref{fig8:stochC}{(B)} to give bimodality. For lower $\n{[c]_a}$,
	meaning larger distance between the first and second fixed point,
	almost no switching was observed. Of course, switching is also a matter of the
	waiting time and stochastic effects: if one waits for long enough, it
	should eventually occur. However, switching times of more than several hours
	are not at the center of this investigation and would mean that switching is
	extremely unlikely. There are two aspects that become relevant in this
	context that we neglect in our analysis but briefly mention here:
	transcription and translation bursts lead to higher stochasticity and cell
	division leads to some discontinuity in the process. The effect of bursts is addressed in~\nameref{S1_Text}.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig9_evolution.eps}
		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
			(A) Shown in blue is the result from one run of the stochastic simulation, and in orange the trajectory obtained from solving the deterministic ODE's. The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
			at an mRNA, protein and intracellular xanthosine count of 0. The parameters that were used are the same as in
			Table~\ref{table1:nondim}, the only exception being the
			extracellular xanthosine concentration, which was chosen to be
			$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
			$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$) just as in
			Fig~\subref{fig8:stochC}{(C)}.
			(B) In blue we plot the waiting time distribution for 1000
			runs of the simulation (same conditions as (A)) to reach
			the steady-state mean.
			We define this time as the elapsed time when the protein
			copy number first reaches 90\% of its value at the upper
			fixed point.
			To better visualize the bulk of the distribution, we
			excluded from the plot $\sim10$ runs with adaptation times
			larger than $6 \cdot 10^5 \unit{s}$.
			The green dashed line indicates the mean of the blue distribution.
			The orange line shows the corresponding deterministic time.
			}
		\label{fig9:StochT}
	\end{figure}
